---
title: "Story2 - API data gathering"
date: "2023-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(httr)
library(jsonlite)
library(tidyverse)
api_key <- .rs.askForPassword("Please input FRED API Key:")
```

## Objective

The Federal Reserve's mandate from Congress is to control inflation and to maintain low unemployment. These seem to be contradictory objectives.  
  
Has the FED been able to fulfill the mandate given to it by Congress?

Remember, the FED raises rate after reviewing the CPI and other data and unemployment (layoffs) occur after company operating costs go up.

## Gathering Data

### Source 1: The Consumer Price Index (CPI)(Bureau of Labor Statistics)

```{r, GET CPI Data}
api_id = 'CPIAUCSL' #set data id to CPI
url <- paste0('https://api.stlouisfed.org/fred/series/observations?series_id=', 
                    api_id, 
                    '&api_key=', 
                    api_key,
                    '&file_type=json') #construct api url

#Make API call with API parameters
res = GET(url,
           query=list(observation_start = "1998-09-20",
                      observation_end = "2023-09-19",
                      frequency = 'm',
                      sort_order = 'asc'))

#retrieve data from API response
data = fromJSON(rawToChar(res$content))

#retrieve the time series from the data in dataframe type
df_cpi <- data$observations |> 
  rename("value_cpi" = "value") |>
  select(-c(realtime_start, realtime_end))
```

### Source 2: The FED Funds Rate (FRED)(Federal Reserve Board)

```{r, GET FED Funds RATE Data}
api_id = 'FEDFUNDS' #set data id to effective federal funds rate
url <- paste0('https://api.stlouisfed.org/fred/series/observations?series_id=', 
                    api_id, 
                    '&api_key=', 
                    api_key,
                    '&file_type=json') #construct api url

#Make API call with API parameters
res = GET(url,
           query=list(observation_start = "1998-09-20",
                      observation_end = "2023-09-19",
                      frequency = 'm',
                      sort_order = 'asc'))

#retrieve data from API response
data = fromJSON(rawToChar(res$content))

#retrieve the time series from the data in dataframe type
df_ffer <- data$observations |> 
  rename("value_ffer" = "value") |>
  select(-c(realtime_start, realtime_end))
```


### Source 3: Unemployment Rate (Bureau of Labor Statistics)

```{r, GET Unemployment Rate Data}
api_id = 'UNRATE' #set data id to effective federal funds rate
url <- paste0('https://api.stlouisfed.org/fred/series/observations?series_id=', 
                    api_id, 
                    '&api_key=', 
                    api_key,
                    '&file_type=json') #construct api url

#Make API call with API parameters
res = GET(url,
           query=list(observation_start = "1998-09-20",
                      observation_end = "2023-09-19",
                      frequency = 'm',
                      sort_order = 'asc'))

#retrieve data from API response
data = fromJSON(rawToChar(res$content))

#retrieve the time series from the data in dataframe type
df_unemp <- data$observations |> 
  rename("value_unemp" = "value") |>
  select(-c(realtime_start, realtime_end))

```


### Data Wrangling
```{r}
df_list = list(df_cpi, df_ffer, df_unemp)
df <- df_list |> 
  reduce(inner_join, by='date')

#export data to csv
write.csv(df, "fred_data.csv", row.names=FALSE)

head(df, 10)
```
